---
title: "Marine Engine Preventive Maintenance Analysis"
author: "Olivier Manthey Salma Fourati Axel Lo Schiavo"
date: "Mai 2025"
format:
  html:
    css: "styles.css"
    self-contained: true
    code-fold: true
    code-summary: "Code"
    warning: false
    message: false
    toc: true
    toc-location: right
    toc-depth: 3
editor: visual
---

# Abstact

Summary of the paper including the main results.

# Introduction

Context and objective of the study

# Literature review

State of the art, previous works, and appropriate citations

# Methodology

Methods and models that are used in the paper

# Data Preparation

## Import the dataset

```{r}
#| message: false
#| error: false 
# Use conda environement for the analysis
library(reticulate)
use_condaenv('MLBA')
```

```{r}
#| message: false
#| error: false 

library(readr)

# Open the marine engine dataset
marine_db <- read_csv("marine_engine_data.csv")
```

## Data cleaning
The raw dataset often contains inconsistencies, missing values, or formatting issues that can negatively impact model performance. In this section, we detail the steps taken to preprocess and clean the data, ensuring it is suitable for subsequent analysis and modeling tasks.
```{r}
#| message: false
#| error: false

library(dplyr)

```

# EDA

# Supervised Learning Analysis

## Data Splitting

## Suport Vector Machine

## Random Forest

## Metrics Analysis

### Tuning of Hyperparameters

# Unsupervised Learning Analysis

## Dimension Reduction

# Results

# Conclusion

# References
