---
title: "Marine Engine Preventive Maintenance Analysis"
author: "Olivier Manthey Salma Fourati Axel Lo Schiavo"
date: "Mai 2025"
format:
  html:
    css: "styles.css"
    self-contained: true
    code-fold: true
    code-summary: "Code"
    warning: false
    message: false
    toc: true
    toc-location: right
    toc-depth: 3
editor: visual
---

# Abstract

Summary of the paper including the main results.

# 1. Introduction

Context and objective of the study

## 1.1 Overview and Motivation

------------------------------------------------------------------------

## 1.2 Data

------------------------------------------------------------------------

## 1.3 Related Work

------------------------------------------------------------------------

## 1.4 Research Question

------------------------------------------------------------------------

# 2. Data

State of the art, previous works, and appropriate citations

## 2.1 Sources

------------------------------------------------------------------------

As already introduced, our research utilizes the “Marine Engines” database owned by *Fijabi J. Adekunle* and hosted on Kaggle, with simulated records from 50 marine engines totalising 5200 instances. This dataset, provided in CSV format, will be used for our analysis aimed at modeling maintenance status for marine engines.

## 2.2 Description

------------------------------------------------------------------------

[timestamp]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Records the weekly time of each observation.

[engine_id]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Unique identifier assigned to each marine engine.

[engine_temp]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Operating temperature of the engine, measured in degrees Celsius.

[oil_pressure]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Pressure of the engine's lubricating oil, in bars.

[fuel_consumption]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Total fuel consumed over the weekly observation period.

[vibration_level]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Level of mechanical vibration detected in the engine.

[rpm]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Revolutions per minute, indicating engine speed.

[engine_load]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Percentage representing how much of the engine's capacity is utilized.

[coolant_temp]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Temperature of the engine coolant fluid.

[exhaust_temp]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Temperature of the exhaust gases emitted by the engine.

[running_period]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Duration (in hours) that the engine was operational.

[fuel_consumption_per_hour]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Average hourly fuel consumption.

[maintenance_status]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Indicates whether the engine requires maintenance, is in normal condition, or is critical.

[failure_mode]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Categorical variable describing the cause of engine failure.

[engine_type]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Classification of the engine (e.g., 2-stroke, 4-stroke, high-speed, low-speed).

[fuel_type]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Type of fuel used by the engine (e.g., Diesel).

[manufacturer]{style="background-color:#f3f0ff; color:#7b2cbf; font-weight:bold; padding:2px 6px; border-radius:4px; font-family:monospace;"}: Manufacturer of the engine (e.g., MAN B&W, Wärtsilä).

## 2.3 Data Preparation

------------------------------------------------------------------------

### 2.3.1 Import the dataset

```{r}
#| message: false
#| error: false 
# Use conda environement for the analysis
library(reticulate)
use_condaenv('MLBA')
```

```{r}
#| message: false
#| error: false 

library(readr)

# Open the marine engine dataset
marine_db <- read_csv("marine_engine_data.csv")
```

### 2.3.2 Data cleaning

The raw dataset often contains inconsistencies, missing values, or formatting issues that can negatively impact model performance. In this section, we detail the steps taken to preprocess and clean the data, ensuring it is suitable for subsequent analysis and modeling tasks.

```{r}
#| message: false
#| error: false

library(dplyr)

# Convert "timestamp" column to datetime
marine_db <- marine_db %>%
  mutate(timestamp = ymd(timestamp))

# Count missing values per column
missing_value <- marine_db %>%
  summaries(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), name_to = "Column", value_to = "Missing_count") %>%
  arrange(desc(Missing_count))

# Display the result
missing_value
```

# 3. **Exploratory Data Analysis (EDA)**

```{r}
#| message: false
#| error: false

library(liver)
library(summarytools)
library(dplyr)
library(psych)
library(ggplot2)
library(ggmosaic)
library(GGally)
```

## **3.1 Data Distribution of Predictive Features (X)**

------------------------------------------------------------------------

## **3.2 Correlation Heatmap: Relationship Between Variables of Interest**

------------------------------------------------------------------------

## **3.3 Target Variables (Y) vs. Predictive Features (X)**

------------------------------------------------------------------------

## **3.4 Relationship between variables of interest**

------------------------------------------------------------------------

# **4. Unsupervised Learning**

# 5. Methodology

Methods and models that are used in the paper

## 5.1 Supervised Learning Analysis

------------------------------------------------------------------------

### 5.1.1 Suport Vector Machine

### 5.1.2 Random Forest

### 5.1.3 Metrics Analysis

### 5.1.4 Tuning of Hyperparameters

# 6. Results

# 7. Conclusion

# \# 8. References

